Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-03-23 rmd/manuscript.Rmd
edit: address Jacques' comments

---
title: "Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems"
output: 
  word_document:
    reference_docx: "word_format.docx"
  #pdf_document:
  #  latex_engine: xelatex
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
csl: https://www.zotero.org/styles/ecosphere
mainfont: Times New Roman
header-includes:
  \usepackage{lineno, textgreek}
  \linenumbers
---

@@

Understanding the drivers of food chain length in natural communities has intrigued ecologists since Elton publicized 'food cycles' in the early 20th century. Proposed drivers of food chain length have included productivity, disturbance regime, ecosystem size, and trophic omnivory. However, current theories have largely assumed simple, two-dimensional habitat architectures and may not be adequate to predict food chain length in ecosystems with a complex, branching structure. Here, we develop a spatially explicit theoretical model which provides an integrated framework for understanding variation in food chain length in branching networks. We show independent, positive influences of ecosystem size and complexity on food chain length. However, these effects were contingent upon other drivers. In particular, the effects of ecosystem size and complexity appeared more clearly in high-disturbance and high-productivity regimes. Our results suggest that ecosystem complexity is an important yet overlooked driver of food chain length in branching ecosystems.

@@

No data were used or produced in this research. R functions for simulations are provided as the R package mcbrnet version 1.3.0 (available at <https://github.com/aterui/mcbrnet>). Codes for simulations, statistical analysis, figures, and tables are available at <https://github.com/aterui/public-proj_fcl-river-theory>.

@@

The origin of variation in food chain length (FCL) has intrigued ecologists for decades due to its importance in regulating top-down effects [@paceTrophicCascadesRevealed1999], energy flow [@wangBiodiversityEcosystemFunctioning2018], and contaminant concentrations in top predators [@kiddEffectsTrophicPosition1998]. Hypotheses regarding controls of FCL can be organized into three broad categories. First, the resource availability hypothesis states that basal productivity controls FCL because imperfect energy conversion through predator-prey interactions restricts energy available to higher trophic levels [@oksanenExploitationEcosystemsGradients1981]. Second, the disturbance hypothesis (or the dynamical stability hypothesis) predicts longer food chains are vulnerable to environmental perturbations (i.e., the dynamical constraint). Thus, habitats with frequent and/or severe disturbance should support fewer trophic levels [@pimmNumberTrophicLevels1977]. Lastly, the ecosystem size hypothesis posits that larger ecosystems should harbor longer food chains [@takimotoEffectsProductivityDisturbance2012; @teruiSpatialDisturbanceSynchrony2019; @postEcosystemSizeDetermines2000].

@@

More recently, food web ecologists have begun to recognize that FCL rarely responds to a single driver, but instead is determined by interactions of multiple factors. For example, @wardMechanisticTheoryAquatic2017 showed that FCL was more responsive to increasing productivity in large ecosystems than small ones, whereas FCL was more sensitive to size in low-productivity systems. Such context-dependent responses to extrinsic factors seem to be tightly coupled with the arrangement of trophic links within a food web. For example, omnivory (feeding at multiple trophic levels) is a strong intrinsic driver of FCL [@takimotoEffectsProductivityDisturbance2012; @postProximateStructuralMechanisms2007; @woottonOmnivoryStabilityFreshwater2017]. When omnivory is absent in three-species communities, the trophic structure or "motif" assumes a simple linear food chain with a maximum trophic position of three (**Figure 1D**). However, FCL decreases with increasing omnivory as the top predator eats lower in the food web [@wardMechanisticTheoryAquatic2017; @woottonOmnivoryStabilityFreshwater2017]. Theoretical work predicts that food webs are destabilized when omnivory is too strong, leading to the loss of species and shorter food chains [@woottonOmnivoryStabilityFreshwater2017]. Conversely, weak omnivory stabilizes the food web and increases the persistence of species within a community, particularly at intermediate levels of productivity and high disturbance levels [@woottonOmnivoryStabilityFreshwater2017; @woottonManyWeakInteractions2016]. Due to the context-dependency of FCL to multiple drivers and the potential for interactive effects, a unified framework of multiple FCL drivers is needed to better understand when and where FCL will respond to these extrinsic and intrinsic factors.

@@

Ample evidence suggests that branching structure controls environmental heterogeneity and dispersal pathways for organisms within river ecosystems [@altermattDiversityRiverineMetacommunities2013; @tonkinRoleDispersalRiver2018]. Water and materials propagate downstream as small streams hierarchically join to form larger streams and rivers. Consequently, environmental signals (e.g., flood disturbance) are highly variable among tributaries while showing a strong within-tributary correlation [@nakamuraDisturbanceRegimesStream2000; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. Confluences are merging points where tributary dynamics aggregate to form distinct environmental conditions in downstream habitat patches. Hence, habitat heterogeneity increases with branching complexity (**Figure 1A**), which we define here as the probability of branching per unit river distance. Highly-branched river networks provide diverse habitats that buffer the impact of large-scale environmental fluctuations, broadening opportunities for species to recolonize habitat patches post-disturbance [@teruiMetapopulationStabilityBranching2018]. Thus, ecosystem complexity in river networks may be a widespread yet overlooked extrinsic environmental driver of FCL.

We simulate trophic community dynamics within branching habitats to investigate the effects of extrinsic controls of FCL, such as ecosystem geometry [size (number of patches) and complexity (branching probability] and disturbance regime. In addition, our simulation model incorporates intrinsic drivers of FCL, including the degree of omnivory, productivity level, and species' dispersal ability, allowing us to assess the context dependency on drivers of FCL across extrinsic factors. We find that ecosystem size and complexity are both positively related to FCL, but the form of this relationship is dependent on other extrinsic and intrinsic factors. Our model is an important step in unifying drivers of FCL across extrinsic and intrinsic drivers and parsing their context-dependency in spatially complex ecosystems.

@@

In our simulation, a disturbance occurs simultaneously across habitat patches with probability $p_m$, but the patches vary in their vulnerability. We simulated the spatial distribution of disturbance intensity (i.e., proportional mortality in the event of stochastic regional disturbance) as follows. A disturbance value, $m_{h}$, was assigned randomly to upstream terminal patch $h$ as $\text{logit}~m_{h} \sim \text{Normal}(\text{logit}~\mu_m, \sigma_{h}^2)$, where $\mu_m$ is the average of disturbance intensity at headwaters in an ordinary scale. The disturbance values propagate downstream through a spatial autoregressive process as:

@@

We obtained the global FCL $\bar{\tau}$ for the last `r unique(df_sim$n_timestep)` time steps. We also calculated the proportion of the patch state ("no species", "B", "B + C", "B + P", and "B + C + P") to infer mechanisms that drive changes in FCL. All analysis was performed in R 4.2.1 [@rcoreteamLanguageEnvironmentStatistical2021]

@@

There were two consistent patterns in the relationship between FCL and ecosystem geometry (size and complexity). First, the effects of ecosystem size and complexity were modulated by disturbance. Second, productivity strengthened the relationship between FCL and ecosystem geometry with important direct effects that were motif dependent. Third, dispersal had a minimal but non-negligible positive effect on FCL. Below we describe these results in detail.

@@

While empirical studies rarely looked at the influence of ecosystem complexity on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled primarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. More empirical research is needed to confirm our theoretical predictions.

@@

Another interesting result of our simulation is the interaction with omnivory. In particular, with no disturbance and low productivity, the FCL of the chain motif had contrasting relationships with ecosystem size (positive) and complexity (negative). It is evident from the one-patch simulation that the chain motif is highly sensitive to local carrying capacity (**Figures S1 and S2**); therefore, FCL followed the changes in the total carrying capacity along the gradients of ecosystem size and complexity. These relationships diminished in omnivory motifs as the dependence of FCL on local carrying capacity weakened or reversed (**Figures S1 and S2**). Importantly, however, disturbance overwhelmed the role of omnivory in modulating the FCL-geometry relationships: as the frequency of intense disturbance increases, the positive relationships between FCL and ecosystem geometry emerged regardless of omnivory levels. This result could explain the consistent support for the ecosystem size hypothesis in various ecosystems [@takimotoEnvironmentalDeterminantsFoodchain2013] and raises the possibility of ecosystem complexity as an important control of FCL in branching networks. Our results also imply that if disturbance is strong and frequent, the effects of ecosystem geometry are robust to food web reorganization, which is often associated with species extirpations or introductions. The strength of omnivory in these novel communities can be increased or decreased, depending on the introduced species and the extant species pool or community. For example, non-indigenous fishes in the Eastern Mediterranean Sea outcompete native species for high-quality prey items, causing native species to increase their consumption of non-preferred prey items [@fanelliDepictingNovelEastern2015]. Yet, such substantial food web reorganization may not alter qualitative relationships between FCL and ecosystem geometry.

>>>>

---
title: "Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems"
output: 
  word_document:
    reference_docx: "word_format.docx"
  # pdf_document:
  #  latex_engine: xelatex
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
csl: https://www.zotero.org/styles/ecosphere
mainfont: Times New Roman
header-includes:
  \usepackage{lineno, textgreek}
  \linenumbers
---

@@

Understanding the drivers of food chain length in natural communities has intrigued ecologists since Elton publicized 'food cycles' in the early 20th century. Proposed drivers of food chain length have included productivity, disturbance regime, ecosystem size, and trophic omnivory. However, current theories have largely assumed simple, two-dimensional habitat architectures and may not be adequate to predict food chain length in ecosystems with a complex, branching structure. Here, we develop a spatially explicit theoretical model which provides an integrated framework for understanding variation in food chain length in branching networks. We show independent, positive influences of ecosystem size and complexity on food chain length. However, the effects of ecosystem size and complexity were contingent upon other factors, appearing more clearly in high-disturbance and high-productivity regimes. Our results suggest that ecosystem complexity is an important yet overlooked driver of food chain length that may increase the resilience to anthropogenic environmental changes.

@@

No primary data were used or produced in this research. R functions for simulations are provided as the R package mcbrnet version 1.3.0 (available at <https://github.com/aterui/mcbrnet>). Codes for simulations, statistical analysis, figures, and tables are available at <https://github.com/aterui/public-proj_fcl-river-theory>.

@@

The origin of variation in food chain length (FCL) has intrigued ecologists for many decades due to its importance in regulating top-down effects [@paceTrophicCascadesRevealed1999], energy flow [@wangBiodiversityEcosystemFunctioning2018], and contaminant concentrations in top predators [@kiddEffectsTrophicPosition1998]. Hypotheses regarding controls of FCL can be organized into three broad categories. First, the resource availability hypothesis states that basal productivity controls FCL because imperfect energy conversion through predator-prey interactions restricts energy available to higher trophic levels [@oksanenExploitationEcosystemsGradients1981]. Second, the disturbance hypothesis (or the dynamical stability hypothesis) predicts longer food chains are vulnerable to environmental perturbations (i.e., the dynamical constraint). Thus, habitats with frequent and/or severe disturbance should support fewer trophic levels [@pimmNumberTrophicLevels1977]. Lastly, the ecosystem size hypothesis posits that larger ecosystems should harbor longer food chains [@takimotoEffectsProductivityDisturbance2012; @teruiSpatialDisturbanceSynchrony2019; @postEcosystemSizeDetermines2000].

@@

More recently, food web ecologists have begun to recognize that FCL rarely responds to a single driver, but instead is determined by interactions of extrinsic (external environmental conditions) and intrinsic factors (species and/or food web attributes). For example, @wardMechanisticTheoryAquatic2017 showed that FCL was more responsive to increasing productivity in large ecosystems than small ones, whereas FCL was more sensitive to size in low-productivity systems. Such context-dependent responses to extrinsic factors seem to be tightly coupled with the arrangement of trophic links within a food web. For example, omnivory (feeding at multiple trophic levels) is a strong intrinsic driver of FCL [@takimotoEffectsProductivityDisturbance2012; @postProximateStructuralMechanisms2007; @woottonOmnivoryStabilityFreshwater2017]. When omnivory is absent in three-species communities, the trophic structure or "motif" assumes a simple linear food chain with a maximum trophic position of three (**Figure 1D**). However, FCL decreases with increasing omnivory as the top predator eats lower in the food web [@wardMechanisticTheoryAquatic2017; @woottonOmnivoryStabilityFreshwater2017]. Theoretical work predicts that food webs are destabilized when omnivory is too strong, leading to the loss of species and shorter food chains [@woottonOmnivoryStabilityFreshwater2017]. Conversely, weak omnivory stabilizes the food web and increases the persistence of species within a community, particularly at intermediate levels of productivity and high disturbance levels [@woottonOmnivoryStabilityFreshwater2017; @woottonManyWeakInteractions2016]. Due to the context-dependency of FCL to multiple drivers and the potential for interactive effects, a unified framework of multiple FCL drivers is needed to better understand when and where FCL will respond to these extrinsic and intrinsic factors.

@@

Ample evidence suggests that branching structure controls environmental heterogeneity and dispersal pathways for organisms within river ecosystems [@altermattDiversityRiverineMetacommunities2013; @tonkinRoleDispersalRiver2018]. Water and materials propagate downstream as small streams hierarchically join to form larger streams and rivers. Consequently, environmental signals (e.g., flood disturbance) are highly variable among tributaries while showing a strong within-tributary correlation [@nakamuraDisturbanceRegimesStream2000; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. Confluences are merging points where tributary dynamics aggregate to form distinct environmental conditions in downstream habitat patches. Hence, habitat heterogeneity increases with branching probability (**Figure 1A**), a proxy for ecosystem complexity in rivers [@teruiMetapopulationStabilityBranching2018; @teruiEmergentDualScaling2021]. Highly-branched river networks provide diverse habitats that buffer the impact of large-scale environmental fluctuations, broadening opportunities for species to recolonize habitat patches post-disturbance [@teruiMetapopulationStabilityBranching2018]. Thus, ecosystem complexity in river networks may be a widespread yet overlooked extrinsic environmental driver of FCL.

We simulate trophic community dynamics within branching habitats to investigate the effects of extrinsic controls of FCL, such as ecosystem geometry [size (number of patches) and complexity (branching probability] and disturbance regime. In addition, our simulation model incorporates established intrinsic drivers of FCL, including the degree of omnivory, productivity level, and species' dispersal ability, allowing us to assess the context dependency on drivers of FCL. We find that ecosystem size and complexity are both positively related to FCL, but the form of this relationship is dependent on other extrinsic and intrinsic factors. Our model is an important step in unifying drivers of FCL across extrinsic and intrinsic drivers and parsing their context-dependency in spatially complex ecosystems.

@@

In our simulation, a disturbance occurs simultaneously across habitat patches with probability $p_m$, but disturbance intensity varies across patches as described below. We simulated the spatial distribution of disturbance intensity (i.e., proportional mortality in the event of stochastic regional disturbance) as follows. A disturbance value, $m_{h}$, was assigned randomly to upstream terminal patch $h$ as $\text{logit}~m_{h} \sim \text{Normal}(\text{logit}~\mu_m, \sigma_{h}^2)$, where $\mu_m$ is the average of disturbance intensity at headwaters in an ordinary scale. The disturbance values propagate downstream through a spatial autoregressive process as:

@@

We obtained the global FCL $\bar{\tau}$ for the last `r unique(df_sim$n_timestep)` time steps. We also calculated the proportion of the patch state ("No species", "B", "B + C", "B + P", and "B + C + P") to infer mechanisms that drive changes in FCL. All analysis was performed in R 4.2.1 [@rcoreteamLanguageEnvironmentStatistical2021]

@@

There were three consistent patterns in the relationship between FCL and ecosystem geometry (size and complexity). First, the effects of ecosystem size and complexity were modulated by disturbance. Second, productivity strengthened the relationship between FCL and ecosystem geometry with important direct effects that were motif dependent. Third, dispersal had a minimal but non-negligible positive effect on FCL. Below we describe these results in detail.

@@

While empirical studies have rarely examined the influence of ecosystem complexity on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled primarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. More empirical research is needed to confirm our theoretical predictions.

@@

Another interesting result of our simulation was the interactive influence of omnivory. In particular, with no disturbance and low productivity, the FCL of the chain motif had contrasting relationships with ecosystem size (positive) and complexity (negative). It is evident from the one-patch simulation that the chain motif is highly sensitive to local carrying capacity (**Figures S1 and S2**); therefore, FCL followed the changes in the total carrying capacity along the gradients of ecosystem size and complexity. These relationships diminished in omnivory motifs as the dependence of FCL on local carrying capacity weakened or reversed (**Figures S1 and S2**). Importantly, however, disturbance overwhelmed the role of omnivory in modulating the FCL-geometry relationships: as the frequency of intense disturbance increases, the positive relationships between FCL and ecosystem geometry emerged regardless of omnivory levels. This result could explain the consistent support for the ecosystem size hypothesis in various ecosystems [@takimotoEnvironmentalDeterminantsFoodchain2013] and raises the possibility of ecosystem complexity as an important control of FCL in branching networks. Our results also imply that if disturbance is strong and frequent, the effects of ecosystem geometry are robust to food web reorganization, which is often associated with species extirpations or introductions. The strength of omnivory in these novel communities can be increased or decreased, depending on the introduced species and the extant species pool or community. For example, non-indigenous fishes in the Eastern Mediterranean Sea outcompete native species for high-quality prey items, causing native species to increase their consumption of non-preferred prey items [@fanelliDepictingNovelEastern2015]. Yet, such substantial food web reorganization may not alter qualitative relationships between FCL and ecosystem geometry.
